Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-20 paper/ms.tex
Editing section describing our version of GenRP

\include{vc}
\usepackage{microtype}

@@

However, this drawback turns out to not be significant:  the exponential kernel
can have a complex exponential coefficients (plus their complex-conjugates), and 
thus approximate an arbitrary autocorrelation function with enough terms, including
quasi-periodic kernels.  It turns out that the sum of Lorentzian components is
a {\it very} good approximation of stellar variability spectra:  the granulation
noise due to stellar convection may be expressed in terms of zero-frequency
Lorentzians, the asteroseismic oscillations may be expressed as non-zero frequency
Lorentizians, as can stellar rotation.  In this paper we extend the \texttt{GenRP} 
approach to kernels of the form:
\begin{equation}
K(\tau) = a_0 \delta(\tau) + \sum_j a_{j=1}^{N_{kernel}} e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
and zero otherwise.  We show that
the extended matrix that embeds this multi-component kernel can be written
in real notation, which speeds up the evaluation of the likelihood function,
and allows optimization software to use automatic differential in computing
derivatives of the likelihood function.

@@

We express the kernel as the sum of exponential functions...

@@

\appendix

>>>>

\include{vc}
\usepackage{microtype}
\usepackage{amsmath}

@@

However, this drawback turns out to not be significant:  the exponential kernel
can have a complex exponential coefficients (plus their complex-conjugates), and 
thus approximate an arbitrary autocorrelation function with enough terms, including
quasi-periodic kernels.  It turns out that the sum of Lorentzian components is
a {\it very} good approximation of stellar variability spectra:  the granulation
noise due to stellar convection may be expressed in terms of zero-frequency
Lorentzians, the asteroseismic oscillations may be expressed as non-zero frequency
Lorentizians, as can stellar rotation.  In this paper we extend the \texttt{GenRP} 
approach to kernels with complex exponentials, which when combined with the
complex conjugate leads to damped harmonic auto-correlation functions which
are flexible enough to describe a wide range of variability, and may still
be solved in $O(N_{time}$ operations.

We show that
the extended matrix that embeds this multi-component kernel can be written
in real notation, which speeds up the evaluation of the likelihood function,
and allows optimization software to use automatic differential in computing
derivatives of the likelihood function.

@@

The kernel is modeled as a sum of damped harmonics,
\begin{equation}
K(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
and zero otherwise.

Now, the damped harmonic kernel components may be rewritten as:
\begin{equation}
 a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
\end{equation}
where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran2015}.
The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
but otherwise the mathematical form of the extended matrix is identical, but with twice as many
exponential kernel components as the real case.

Although this is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
arithmetic.  In addition, it turns out that the complex conjugate component, $\beta^* = \beta_R - i\beta_I$, in 
the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
component $\beta = \beta_R + i \beta_I$;  thus the computation is redundant.  Finally, complex
arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
which is what we would like to use to compute the derivative of the likelihood (as it is very difficult
to compute this explicitly!).

Consequently, we use the fact that $\exp{i\beta_I \tau} =  \cos{\beta_I \tau} + i \sin{\beta_I \tau}$ to
rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
may be replaced by a $2\times 2$ matrix, $ \begin{pmatrix} 0 & 1 \\ -1 & 0\end{pmatrix}$, so that
$\cos{\beta_I \tau} + i \sin{\beta_I \tau}$ may be rewritten as the rotation matrix
$\begin{pmatrix} \cos{\beta_I \tau} & \sin{\beta_I \tau} \\ -\sin{\beta_I \tau} & \cos{\beta_I \tau}\end{pmatrix}$.
We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex 
equations, resulting in a real extended matrix that is the same size as the complex matrix.
This transformation leads to about a factor of two in computational savings, but has the
advantage of allowing the automatic differentiation due to the use of real arithmetic.

\subsection{Kernel components}

@@

\appendix \label{appendixa}
